The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-05-07 doc/SA-Hermaphrodites-wDemography.tex
Updated README.md
Self explanatory commit.
empty commit for fixMaster
Again. empty commit to fix erroneous merge
Minor edits main text and suppl

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{pdflscape}

@@

\begin{document}
\title{The demographic costs of sexually antagonistic selection in simultaneous hermaphrodites}

@@

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population intrinsic growth rate remains positive. In addition, we analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that as population intrinsic growth rates approach one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M. guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take the population dynamical consequences of SA selection into account, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \citep{kokko2003sexy}, but few papers since then explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019b} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \citep{deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which balanced SA polymorphisms are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019b} under obligate outcrossing. A full derivation of the model and analyses is presented in the \hl{Online Supplementary Material}, and all computer code necessary to reproduce the results are available at \url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. In the \hl{Online Supplement}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}).

@@

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of extant individuals produced by selfing or outcrossing ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrix for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}). Further details of the derivations of $\mcal{U}^S$, $\mcal{U}^X$ ,  $\mcal{F}^S$, and $\mcal{F}^X$ are provided in the \hl{Online Supplement}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[htbp]
\centering
\caption{\bf Definition of terms.}
\label{tab:Terms}
\begin{tabular}{ l p{0.7\linewidth} r }
 \toprule
Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ \\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mbb{F}^{\prime}$ & Block diagonal fertility matrix & $\omega g \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$ \\
$\mbf{H}^{S}_{i}$,\, $\mbf{H}^{X}_{i}$ & Parent-offspring genotype maps & $g \times g$ \\
$\mbf{W}$   & Allele segregation matrix & $\omega \times g$ \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times g$ \\

@@

The SA locus does not affect survival and transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\sigma$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_a
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

Throughout our analyses, we distinguish between early- and late-acting inbreeding depression. We quantify early-acting inbreeding depression using $\delta$, and late-acting inbreeding depression using $\delta_i$ (where $i \in \{j,a,\gamma\}$). $\delta$ denotes the fraction of self-fertilized ovules that do not develop into juveniles due to inbreeding depression. An important difference between early- and late-acting inbreeding depression in the model is that $\delta$ does not affect $\mbf{U}^S$, the survival matrix for offspring produced by selfing (see Eq.~\ref{eq:Atilde}), while the $\delta_i$ terms directly alter the stage-specific survival and transition rates of offspring produced by selfing (see Eq.~\ref{eq:US}).

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we present results for scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of additive SA fitness effects ($h = 1/2$) and dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$). We relax these assumptions to explore the consequences of sex-specific dominance, which may evolve under asymmetric SA selection \citep{SpencerPriest2016}, on our model predictions in \hl{Appendix X}).

@@

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized the model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic matrix data available on the public demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix C}.

@@

Using these combined demographic rates, selfing rates, and inbreeding depression estimates, we constructed a corresponding stage $\times$ genotype mendelian matrix model with a single SA locus affecting female and male fertility (as described above). With the empirically parameterized model, we are able to make predictions about the scope for demographically viable SA polymorphism at a single locus in hypothetical populations with the same demographic rates as observed in \citet{PetersonEtAl2016}, for a range of selfing and inbreeding depression rates observed in other natural populations. We stress, however, that these are illustrative rather than explicit predictions of the likelihood of SA polymorphism in any specific population, and ignore measurement error for the estimated demographic rates.

@@

When using demographic rates for either the Eagle Meadows or Low-Elevation Perennial populations, inv6 falls within demographically viable polymorphic parameter space (fig.~\ref{fig:mimulusFig}; estimated selection coefficients for inv6 were $(s_f,\,s_m) = (0.30,\, 0.31)$). Whether under obligate outcrossing or partial selfing, inv6 always falls squarely in the middle of SA polymorphic space when using the locally adapted Eagle Meadows demographic rates. In contrast, when the selfing rate is at the higher end of empirical estimates for the Iron Mountain population in which inv6 has been documented ($C = 0.29$), inv6 falls near the upper boundary but still within SA polymorhic space when using the Low-Elevation Perennial demographic rates. This happens because under the empirical estimates of selfing and inbreeding depression, the polymorphic parameter space shifts downwards. Additionally, inv6 falls much closer to the extinction threshold under partial selfing. This suggests that even relatively small perturbations to demgraphic rates or selection coefficients could result in non-locally adapted populations being unable to support the demographic costs associated with segregating SA alleles with selection coefficients of similar magnitude to inv6.

@@

Classic population genetics theory predicts that sexually antagonistic selection is unlikely to maintain genetic variation except under narrow conditions, with polymorphism requiring either finely balanced or unusually strong selection, or partially recessive fitness effects through each sex \citep{Kidwell1977,Pamilo1979,Prout2000, ConnallonClark2014}. Extensions of the theory have identified numerous ways in which the conditions for polymorphism become more permissive in both dioecious and hermaphroditic organisms, including genetic linkage of SA loci, the evolutioin of sex-specific dominance, population subdivision, and life-cycle complexity \citep[e.g.,][]{Patten2010,JordanConnallon2014,JordanCharlesworth2012,SpencerPriest2016, ConnallonSharmaOlito2019, Olito-etal-2018}. However, by ignoring the demographic consequences of SA genetic variation, these population genetic models have missed the possibility that SA polymorphisms may not be viable under realistic parameter conditions, and therefore unlikely to be observed in natural populations. By linking the individual-level fitness consequences of SA selection to population level demographic rates, our theoretical framework provides several key insights into the processes shaping SA genetic variation in natural populations.

@@

Like most demographic matrix models, ours also assumes female demographic dominance, where population growth rates are determined entirely by female fecundity \citep{Caswell2001,CaswellEtAl2018,deVriesCaswell2019b}. Yet, many hermaphroditic populations experience limitation of reproductive success due to both quantity and quality of male gametes \citep[e.g.,][]{Yund2000,AizenHarder2007, Harder2016}. Explicitly modeling male gamete production is a natural extension to our modeling framework, and would enable us to analyze how tension between the demographic consequences of SA fitness variation through both female and male function alters or abolishes asymmetries in the extinction thresholds and polymorphic parameter space \citep[e.g.,][]{Tazzyman2015}.  Interestingly, different forms of self-fertilization can aid in reproductive assurance, and should therefore have demographic consequences -- for example, under pollen limitation and delayed selfing (where only ovules that fail to receive outcross pollen are selfed), the selfing rate will be a function of genotype frequencies because these will directly influence pollen production \citep{HarderBarrett2006}.

Another major simplifying assumption in our model was that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization.

>>>>

\documentclass[11pt,draft]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{pdflscape}

@@

\begin{document}
\title{The demographic costs of sexually antagonistic selection in partially selfing populations}

@@

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the intrinsic population growth rate remains positive. In addition, we analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that as population intrinsic growth rates approach one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M. guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take the population dynamical consequences of SA selection into account, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \citep{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which balanced SA polymorphisms are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing. A full derivation of the model and analyses is presented in the \hl{Online Supplementary Material}, and all computer code necessary to reproduce the results are available at \url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. \textcolor{red}{In the \hl{Online Supplement}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}). Do we???}

@@

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of extant individuals produced by selfing or outcrossing ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrices for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}). Further details of the derivations of $\mcal{U}^S$, $\mcal{U}^X$ ,  $\mcal{F}^S$, and $\mcal{F}^X$ are provided in the \hl{Online Supplement}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[htbp]
\centering
\caption{\bf Definition of terms.}
\label{tab:Terms}
\begin{tabular}{ l p{0.7\linewidth} r }
 \toprule
Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ g\\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$g \\
$\mbf{H}^{S}_{i}$,\, $\mbf{H}^{X}_{i}$ & Parent-offspring genotype maps & $g \times g$ \\
$\mbf{W}$   & Allele segregation matrix & $2 \times \omega $g \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times 2\omega g$ \\

@@

The SA locus does not affect survival and transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\delta$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_a
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

Throughout our analyses, we distinguish between early- and late-acting inbreeding depression. We quantify early-acting inbreeding depression using $\delta$, and late-acting inbreeding depression using $\delta_i$ (where $i \in \{j,a,\gamma\}$). $\delta$ denotes the fraction of self-fertilized ovules that do not develop into juveniles due to inbreeding depression. An important difference between early- and late-acting inbreeding depression in the model is that $\delta$ affects the production of new individuals, whereas the $\delta_i$ affect the demographic rates of extant individuals, contained in $\mbf{U}^S$ (see Eq.~\ref{eq:US}).

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we present results for scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of additive SA fitness effects ($h = 1/2$) and dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$). \textcolor{red}{We relax these assumptions to explore the consequences of sex-specific dominance, which may evolve under asymmetric SA selection \citep{SpencerPriest2016}, on our model predictions in \hl{Appendix X}). }

@@

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized the model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic data available on the public demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix C}.

@@

Using these combined demographic rates, selfing rates, and inbreeding depression estimates, we constructed a corresponding stage $\times$ genotype mendelian matrix model with a single SA locus affecting female and male fertility (as described above). With the empirically parameterized model, we are able to make predictions about the scope for demographically viable SA polymorphism at a single locus in hypothetical populations with the same demographic rates as observed in \citet{PetersonEtAl2016}, for a range of selfing and inbreeding depression rates observed in other natural populations. We stress, however, that these are illustrative rather than explicit predictions of the likelihood of SA polymorphism in any specific population, and they ignore measurement error for the estimated demographic rates.

@@

When using demographic rates for either the Eagle Meadows or Low-Elevation Perennial populations, inv6 falls within demographically viable polymorphic parameter space (fig.~\ref{fig:mimulusFig}; estimated selection coefficients for inv6 were $(s_f,\,s_m) = (0.30,\, 0.31)$). \hl{I like Crispins idea of reporting predicted inv6 frequencies in our model, it nicely illustrates another useful output of the model} Whether under obligate outcrossing or partial selfing, inv6 always falls squarely in the middle of SA polymorphic space when using the locally adapted Eagle Meadows demographic rates. In contrast, when the selfing rate is at the higher end of empirical estimates for the Iron Mountain population in which inv6 has been documented ($C = 0.29$), inv6 falls near the upper boundary but still within SA polymorhic space when using the Low-Elevation Perennial demographic rates. \hl{Another place where we could report inv6 frequencies?} This happens because under the empirical estimates of selfing and inbreeding depression, the polymorphic parameter space shifts downwards. Additionally, inv6 falls much closer to the extinction threshold under partial selfing. This suggests that even relatively small perturbations to demographic rates or selection coefficients could result in non-locally adapted populations being unable to support the demographic costs associated with segregating SA alleles with selection coefficients of similar magnitude to inv6.

@@

Classic population genetics theory predicts that sexually antagonistic selection is unlikely to maintain genetic variation except under narrow conditions, with polymorphism requiring either finely balanced or unusually strong selection, or partially recessive fitness effects through each sex \citep{Kidwell1977,Pamilo1979,Prout2000, ConnallonClark2014}. Extensions of the theory have identified numerous ways in which the conditions for polymorphism become more permissive in both dioecious and hermaphroditic organisms, including genetic linkage of SA loci, the evolutioin of sex-specific dominance, population subdivision, and life-cycle complexity \citep[e.g.,][]{Patten2010,JordanConnallon2014,JordanCharlesworth2012,SpencerPriest2016, ConnallonSharmaOlito2019, Olito-etal-2018}. However, by ignoring the demographic consequences of SA genetic variation, these population genetic models have missed the possibility that SA polymorphisms may not be viable under realistic parameter conditions, and therefore unlikely to be observed in natural populations. By linking the individual-level fitness consequences of SA selection to population level consequences, our theoretical framework provides several key insights into the processes shaping SA genetic variation in natural populations.

@@

Like most demographic matrix models, ours also assumes female demographic dominance, where population growth rates are determined entirely by female fecundity \citep{pollard1975mathematical,Caswell2001,iannelli2005gender}. Yet, many hermaphroditic populations experience limitation of reproductive success due to both quantity and quality of male gametes \citep[e.g.,][]{Yund2000,AizenHarder2007, Harder2016}. Explicitly modeling male gamete production is a natural extension to our modeling framework, and would enable us to analyze how tension between the demographic consequences of SA fitness variation through both female and male function alters or abolishes asymmetries in the extinction thresholds and polymorphic parameter space \citep[e.g.,][]{Tazzyman2015}.  Interestingly, different forms of self-fertilization can aid in reproductive assurance, and should therefore have demographic consequences -- for example, under pollen limitation and delayed selfing (where only ovules that fail to receive outcross pollen are selfed), the selfing rate will be a function of genotype frequencies because these will directly influence pollen production \citep{HarderBarrett2006}.
,
Another major simplifying assumption in our model was that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization.
